File-based localization of numerical perturbations in data analysis pipelines

### 2020-04-05 paper.tex
Fixed Fig 2

% There are different approaches to improve the reproducibility of the analysis, 
% including containerization techniques that encapsulate software/hardware dependencies, 
% provenance capturing tools, and version control systems. 
% However, a comprehensive solution requires to fix the numerical instabilities instead of 
% masking the problem. For this purpose, bagging is one solution that has been used to 
% stabilize motion estimation analysis in fMRI~\cite{Glatard2018hbm}. 
% Bootstrap aggregation reduces the instabilities, but it needs expensive computations.
% As the second solution, a debugging tool can help to identify and 
% fix the instabilities in the pipeline. Therefore, we introduce 
% the numerical reproducibility measurement tool named \emph{Spot} to identify 
% the processes in the pipeline that create numerical differences across different conditions. 
% Furthermore, we can stabilize the identified processes in the next steps.
\todo{explain creation vs propagation}
\todo{define pipeline}

@@

\toolname identifies the processes in a pipeline that produce different
results when executed in two different conditions. First, a directed
bipartite provenance graph is recorded for each subject, where nodes
represent application processes and files, and edges represent read and
write file accesses (Step 1 in Figure~\ref{fig:overview-tool}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified (Step 2) and
disambiguated, resulting in a provenance \emph{DAG} (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Step 3). 
DAGs produced in the two conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Step
4). Finally, a labeled graph is produced that identifies the
non-reproducible processes in the pipeline.

To ensure that a file can be unambiguously associated with the process that
created it, we assume that pipeline processes can be re-ordered such that:
\begin{enumerate}
\item Processes don't run concurrently;
\item Each process sequentially reads, computes, and writes. 
\end{enumerate}
In practice, pipeline processes may still run concurrently provided that
they don't write concurrently to the same files. A process may also
interleave file writes with computing, for instance when different file
blocks are processed sequentially. However, only a single version of the
file must eventually be made available to the other processes. Similarly,
in case processes write and delete temporary files throughout their
execution, these temporary files must not be read by any other process in
the pipeline.

In addition, we also require that processes are
associated to a command line (executable and arguments), to facilitate
process instrumentation.

@@

Provenance graphs are often subject-dependent, due to variations in input data among subjects.
%, which justifies the processing
% of multiple subjects. For instance, data acquisitions may be
% repeated a different number of times in each subject, for noise-reduction
% purposes. 
Some of these differences can be neglected, for instance when a
data decompression step is present at the beginning of the execution for
some subjects only, but other differences cannot, for instance
when complete different processing paths are used in different subjects. \toolname 
includes helpers to identify different instances of provenance graphs.
For this purpose, \toolname measures the distance between process trees using 
the tree edit distance method~\cite{zhang1989simple}, as implemented in 
the zss Python package. This distance is defined as the minimum number of
edit operations to transform one tree into the other. Three edit operations
are considered: node label modification, node removal, and node insertion.
Each operation has an associated cost of 1.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\linewidth}
  \centering
  \includegraphics[width=.7\linewidth]{images/p-graph}
  \caption{Raw provenance graph (\reprozip output), with transient files shown in gray.}
  \label{fig:provenance-graph}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.4\linewidth}
  \centering
  \includegraphics[width=.7\linewidth]{images/p-graph-dag}
  \caption{Provenance DAG, with disambiguated transient files.}
  \label{fig:provenance-dag}
\end{subfigure}
  % \includegraphics[width=.8\columnwidth]{images/provenance-graphs}
  \caption{Provenance graphs
  created from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Processes are represented with circles, files with rectangles, and read/write accesses 
  with plain edges. 
  For visualization purposes, the process tree is represented too, with
  dashed edges. Every node in the graph is labeled using (1) a process id
  created by our reconstruction, (2) the name of the executable run by the
  process or the file name. Process 0 is the initial call to the
  \texttt{sh} interpreter; the other processes are the calls to FSL bet,
  stats and maths made in Listing~\ref{listing:sample-script}. Process 73
  is forked by process 2: it was captured by \reprozip while it did not
  appear in Listing~\ref{listing:sample-script}.
}
  \label{fig:simple_script}
\end{figure}

@@

\subsection{Labelling processes}

After capturing transient files, we re-run the pipeline
step by step in Condition 2 to label processes. The output files
produced by a process in Condition 2 are compared to the ones produced
in Condition 1. If no differences are found, the process is marked as
reproducible. Otherwise, the process is marked as non-reproducible and the
output files produced in Condition 1 are copied to Condition 2, to ensure
that differences do not propagate further in the pipeline. Processes are
instrumented transparently, through a modification of the \texttt{PATH}
variable similar to the one described previously. By default, differences
in output files are identified by comparing file checksums. Other
comparison functions can also be defined for specific file types, for
instance to ignore file headers or file sections containing timestamps.
\toolname finally creates a labeled
provenance graph highlighting non-reproducible processes.

Figure~\ref{fig:iterations} illustrates the incremental labelling 
process for the example in Figure~\ref{fig:simple_script}. 
At step 1, process \texttt{73\#bet2} is labeled as non-reproducible (red) 
as it produces files with differences. Subsequently, the files produced by \texttt{73\#bet2} in  
Condition 2 are replaced with the files produced by \texttt{73\#bet2} in 
Condition 1.
At step 2, process \texttt{75\#fslmaths} is executed and then labeled 
as reproducible (green) as it produced files without differences.
At step 3, \texttt{76\#fslstats} is also labeled as 
reproducible as it didn't produce any file.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{images/labelling-process}
  \caption{Illustration of the labelling approach. Non-reproducible processes are shown in red.}
  \label{fig:iterations}
\end{figure}

@@

We applied the \toolname to find the origin of between-OS differences that
frequently occur in neuroimaging pipelines. In particular, we studied the minimal
pre-processing pipelines released by the Human Connectome Project
(\href{https://www.humanconnectome.org}{HCP}), a leading initiative in
neurosciences.

@@

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{images/pfs_heatmap.png}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the FLIRT tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects.
        }
  \label{fig:pfs_freq}
\end{figure*}

@@

>>>>

% There are different approaches to improve the reproducibility of the analysis, 
% including containerization techniques that encapsulate software/hardware dependencies, 
% provenance capturing tools, and version control systems. 
% However, a comprehensive solution requires to fix the numerical instabilities instead of 
% masking the problem. For this purpose, bagging is one solution that has been used to 
% stabilize motion estimation analysis in fMRI~\cite{Glatard2018hbm}. 
% Bootstrap aggregation reduces the instabilities, but it needs expensive computations.
% As the second solution, a debugging tool can help to identify and 
% fix the instabilities in the pipeline. Therefore, we introduce 
% the numerical reproducibility measurement tool named \emph{Spot} to identify 
% the processes in the pipeline that create numerical differences across different conditions. 
% Furthermore, we can stabilize the identified processes in the next steps.
\todo{explain creation vs proSpagation}
\todo{define pipeline}
\todo{explain the need for reduced instrumentation}

@@

\toolname identifies the processes in a pipeline that produce different
results in different execution conditions. First, a directed
bipartite provenance graph is recorded for each subject, where nodes
represent application processes and files, and edges represent read and
write file accesses (Step 1 in Figure~\ref{fig:overview-tool}). Second,
transient files, i.e., files that are either deleted during pipeline
execution or modified by multiple processes, are identified (Step 2) and
disambiguated, resulting in a provenance DAG (Directed Acyclic
Graph) in which file nodes have a single parent (in-degree of 1) (Step 3). 
DAGs produced in different conditions are then compared, in
a step-by-step execution that prevents the propagation of differences in the pipeline (Step
4). The resulting labeled graph identifies the
non-reproducible processes in the pipeline.

To ensure that a file can be unambiguously associated with the process that
created it, we assume that the pipeline can be transformed such that:
\begin{enumerate}
\item Processes don't run concurrently;
\item Each process sequentially reads, computes, and writes. 
\end{enumerate}
In practice, pipeline processes may still run concurrently provided that
they don't write concurrently to the same files. A process may also
interleave file writes with computing, for instance when different file
blocks are processed sequentially. However, only a single version of the
file must eventually be made available to the other processes. In
particular, in case a process deletes a file that it had created itself,
this file must not be used by any other process. Finally, we also require that processes are
associated to a command line (executable and arguments), to facilitate
process instrumentation.

@@

Provenance graphs are often subject-dependent, due to variations in input data among subjects.
%, which justifies the processing
% of multiple subjects. For instance, data acquisitions may be
% repeated a different number of times in each subject, for noise-reduction
% purposes. 
Some of these differences can be neglected, for instance when a data
decompression step is present at the beginning of the execution for some
subjects only, but other differences cannot, for instance when complete
different processing paths are used in different subjects. \toolname
includes helpers to identify different instances of provenance graphs. In
particular, it can cluster process trees, where nodes are processes and
edges are \texttt{fork()} or \texttt{clone()} system calls, using the tree
edit distance~\cite{zhang1989simple} implemented in Python's \texttt{zss} package.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\linewidth}
  \centering
  \includegraphics[width=.7\linewidth]{images/p-graph.pdf}
  \caption{Raw provenance graph (\reprozip output), with transient files shown in gray.}
  \label{fig:provenance-graph}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.4\linewidth}
  \centering
  \includegraphics[width=.7\linewidth]{images/p-graph-dag.pdf}
  \caption{Provenance DAG, with disambiguated transient files.}
  \label{fig:provenance-dag}
\end{subfigure}
  % \includegraphics[width=.8\columnwidth]{images/provenance-graphs}
  \caption{Provenance graphs
  created from the example pipeline in
  Listing~\ref{listing:sample-script}.
  Processes are represented with circles, files with rectangles, and read/write accesses 
  with plain edges. 
  For visualization purposes, the process tree is represented too, with
  dashed edges. Every node in the graph is labeled using (1) a process id
  created by our reconstruction, (2) the name of the executable run by the
  process or the file name. Process 0 is the initial call to the
  \texttt{sh} interpreter; the other processes are the calls to FSL bet,
  stats and maths made in Listing~\ref{listing:sample-script}. Process 73
  is forked by process 2: it was captured by \reprozip while it did not
  appear in Listing~\ref{listing:sample-script}.
}
  \label{fig:simple_script}
\end{figure}

@@

\subsection{Labeling processes}

After capturing transient files, we re-run the pipeline
step by step in Condition 2 to label processes. The output files
produced by a process in Condition 2 are compared to the ones produced
in Condition 1. If no differences are found, the process is marked as
reproducible. Otherwise, the process is marked as non-reproducible, and the
output files produced in Condition 1 are copied to Condition 2, to ensure
that differences do not propagate further in the pipeline. Processes are
instrumented transparently, through a modification of the \texttt{PATH}
variable similar to the one described previously. By default, differences
in output files are identified by comparing file checksums. Other
comparison functions can also be defined for specific file types, for
instance to ignore file headers or file sections containing timestamps.
\toolname finally creates a labeled
provenance graph highlighting non-reproducible processes.

Figure~\ref{fig:iterations} illustrates a hypothetical incremental labeling 
process for the example in Figure~\ref{fig:simple_script}. 
At step 1, process \texttt{73\#bet2} is labeled as non-reproducible (red) 
as it produces files with differences, and the files produced by \texttt{73\#bet2} in  
Condition 2 are replaced with the files produced by \texttt{73\#bet2} in 
Condition 1. At step 2, process \texttt{75\#fslmaths} is executed and then labeled 
as reproducible (green) as it produced files without differences.
At step 3, \texttt{76\#fslstats} is also labeled as 
reproducible as it didn't produce any file.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{images/p-graph-dag-labelled.pdf}
  \caption{Illustration of the labeling approach. Non-reproducible processes are shown in red.}
  \label{fig:iterations}
\end{figure}

@@

We applied the \toolname to the minimal
pre-processing pipelines released by the Human Connectome Project
(\href{https://www.humanconnectome.org}{HCP}), a leading initiative in
neurosciences.

@@

% make figure in two columns use {figure*}
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{./data/pfs-20sbj-output/heatmap.png}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps (N=20). 
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  number of subjects where the corresponding process appears. For example,
  the FLIRT tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects.
        }
  \label{fig:pfs_freq}
\end{figure*}

@@

Mention non-instrumentation / reprozip
